Binning Strategies and Related Loss for Binned Scatterplots of Large Data

### 2016-12-05 AuthorResponseToReviewers.Rmd

### Responses to Reviewer 1 Comments/Questions

*The authors discuss different binning strategies and the resulting aggregation- induced errors. Both errors, loss of spatial resolution and loss of frequency information are introduced and formalized. Binning is an important preprocessing step and research in this area is highly relevant. However, there are some major points missing, making this work less convincing.*

*The most crucial point is in my opinion the missing discussion of the influence of different data distributions. In the visualization community exists research on different descriptors for two-dimensional data called Scagnostics. When I started reading this article, I expected a thorough discussion of different data distributions/properties and the influence on different binning strategies and error outcomes.*

- We are grateful that the review brought the literature on scagnostics to our attention; the work and scagnostic metrics formalized in the works by Wilkerson, Anand, Grosmann and Wills will be valuable to the discussion of binned scatterplots. Specifically scagnostic literature contributes to Section 5 of our paper in the exploration of loss under a variety of simulated distribution. You will find that a discussion of the relevance of scagnostics has been added immediately preceding the start of subsection 5.1. It is interesting that the striation score is so small, given that the data resolution issue can lead to artificially induced stripping is seen in so prominently in Figure 6.

@@

*I don’t buy the advice to “use between four and seven frequency”. There simply isn’t empirical evidence in the paper to support this claim.*

- We do not want to allow more color bin than we can *easily* differentiate because the our expressed purpose for working with discretized color scales in this scenario is to have made loss of frequency information *entirely* explicit. While working with discretizing shades we found ourselves worrying about our ability to perfectly match bins to a legend after approximately 6 shades. We do not claim to have gathered empirical evidence, rather we defer to Healey and Enn's Theory on discernible colors (typically 4-7) to back out experience. Figure 9 was included for the reader to assess if they felt our suggestion was justifiable. A statement clarifying that the recommendations is to bolster the reliable perceptual mapping has been added to the discussion section.

*From a human-computer interaction perspective, it is also limiting when the paper talks about spatial loss and frequency loss without any mention of user tasks (e.g. comparison, outlier detection). Do these metric matter for all the tasks involved in reading the scatterplots?*

- A statement on the connection of spatially-related user takes has been added to the introduction of spatial loss in section 4.1. Also, the difficult comparison of bin frequency information through continuous shade has been added to the discussion on why frequency binning may be employed in Section 4.2.

### 2016-12-05 versionForJCGS/PaperJCGS.Rnw
final review updates

Binning algorithms used in making distributional approximations can be traced back to Pearson's work with the binomial approximation to the normal, where he mentions the need to define an origin and binwidth for segmenting the normal distribution \citep{Pearson1895}. \ktm{Sturges followed with in early work in formalizing histograms specification \citep{sturges1926choice}.}  More recently Scott has presented discussion on the importance of binning specification in the creation of histograms to appropriately display one dimensional density approximations \citep{scott1979}. \citet{scott1992} extends to the properties of multivariate binning strategies.

@@

When the individual points of a scatterplot are collapsed to bin centers to be displayed as tiles in a binned scatterplot there is a loss of the location information.  This can be expressed as the Euclidean distance between points and the visual center of the tiles (i.e.\ the bin centers). Note that other distance metrics could be used, but the Euclidean distance has a desirable interpretability in $\mathbb{R}^2$. The \textit{total spatial loss}, $L^S$, is defined as 
\begin{eqnarray}\label{totspatloss}
L^S = \sum_{i=1}^{n} L_i^S = \sum_{i=1}^{n} \sqrt{ \left(x_i-b_X(x_i)\right)^2 + \left(y_i-b_Y(y_i)\right)^2 }
\end{eqnarray} 
%
where $L_i^S$ is the loss in the  $i$th observation. 
%
Figure~\ref{fig:spatlossdemo2} visually displays the spatial loss for the data from  Table~\ref{DataReductionExample} as a result of standard rectangular binning. Observations $(x_i,y_i)$ and bin centers $\left(x_\ell^\ast,y_\ell^\ast\right)$ are displayed as black points and gray crosses, respectively. The length of line segments connecting these represent $L^S_i$, the spatial loss for each observation; thus, the combined length of all line segments represents the total spatial loss, $L^S$.  For random assignment the total spatial loss can be calculated by simply replacing the standard binning function with the random binning function in Equation~\ref{totspatloss}.

@@

% It is important to note that there is no spatial loss for the traditional scatterplot, as the points are rendered at exact coordinates. The reason we are willing to sacrifice precision of location information is to alleviate the inherent loss of frequency information from over-plotting, which can be immense for large data sets. When a point is rendered in a traditional scatterplot there is no graphical change once any additional points are placed at the same coordinates; thus visually implying a frequency of one data value at the location. This implicit loss of frequency information is then exacerbated when points are rendered as a circle with a radius large enough to also partially cover nearby points. In binned scatterplots, we have traded the exact locations for exact frequencies within bins. 
% 
Bin counts can be displayed using a continuous shading scale in HSV color space \citep{colorperception} and thus we can theoretically map frequency to shade perfectly. While the tiles of the binned scatterplot would be \textit{rendered} precisely, the ability of a human with average vision to extract that information by visually mapping the tile shade to a frequency scale in the plot legend is largely imprecise. Color perception of shade is \ktm{influenced by the contrasting effect of surrounding shades}, allowing an inaccurate mapping of tile shade to the corresponding shade in the plot legend \ktm{\citep{bartleson1967brightness,adelson1993perceptual,fairchild2013color}}. See Figure~\ref{fig:FreqLossDemoPlot} for a demonstration of context sensitivity of colors. It is therefore not realistic, to expect readers to be able to decode frequency from a continuous shade scheme, even though theoretically we can perceive shades continuously \citep{Leong}.  %Even though there is a lot of spread between the colors in this example, most onlookers have trouble answering the question about the relative relationship of the shading of the darkest/lightest points in each cloud of points.

@@

%Density changes are impossible to make out in the solid black areas of the plots. 97.4\%, 98.7\% and 99.1\% of the points are completely hidden behind at least one other point due to over-plotting in the coarse data for the uniform, normal and exponential simulations, respectively.
%
The \textit{data resolution} is defined as the smallest increment between successive data values. To observe loss from binning under more realistic conditions, we create three data sets by rounding the values from the originally simulated data sets to the nearest even number, hus a data resolution of 2 units in each dimension. This coarse version of the original simulated data is displayed in the bottom row of Figure~\ref{fig:SimDataScatterplots}. By exploring the loss properties for both the \textit{fine} and \textit{coarse} versions of the data, we identify which binning options are robust to the resolution at which data is recorded.

@@

To demonstrate the importance of properly selecting binwidths we consider the coarse version of the simulated bivariate-uniform data which is recorded to a resolution of two units in each dimension \ktm{-- thus even integer sized bins are most desireable. Figure~\ref{fig:UniformStripes}(a) displays the binned scatterplots under several scenarios. Under standard binning we see the white-space gaps with one-by-one unit bins, vertical and horizontal artificial stripes with five-by-five unit bins, and the appropriate view of an evenly spread density with four-by-four unit bins. Figure~\ref{fig:UniformStripes}(b) demonstates that with coarse data resolution there are departures from the linear relationship between spatial loss and bin sizes; with small spikes in spatial loss where the bin size is highly misaligned with the data resolution and slight drops in spatial loss when aligned perfectly. In this we can see that artificial striping is a symptom of poorly aligned bin specification.}

@@

Note that random binning is effective at smoothing out the artificial striping patterns \ktm{when many point fell along bin boundaries -- accomplishing the motiviation for its development. However due to inflated spatial loss due to the stocastic bin assignment, standard binning with properly selected bin dimension is the better option}.

@@

We now revisit the baseball data used earlier and construct binned scatterplots using the loss-based specification principles developed in the previous sections. \ktm{In the spatial binning step we learned three rules of thumb: (1) use smaller bins to achieve lower loss of spatial information, (2) standard binning has less loss than random binning but we need to be careful to choose bin dimensions that align with data resolution, and (3) offset the origin by a factor of half the data resolution to better center point in bins.}

@@

\ktm{Due to imperfect human perception may elect to use frequency binning to discretize the shading scale with the goal to make the loss of frequency information more explicit. If we aim to have highly accurate perceptual mapping,} it is recommended to use at most between four and seven distinct shades. This is done to minimize the frequency loss within the bounds of human perceptual ability to distinguish multiple shades simultaneously. Using quantile frequency binning or standard log frequency binning are shown to be reasonable methods -- with slight differences of interpretability -- for handling situations with heavily skewed bin counts. \ktm{Frequency binning to discretize the shading scale allows us to explicitly quantify the loss of frequency information, however future perception research is needed to quantify the implicit frequency loss in reading binned scatterplots using continuous shading and the impact of contrasting effects.}

Future implementations of software for constructing binned scatterplots would be well served to allow for choices in specification of binning algorithms. The findings of this research provide suggestions for reasonable default settings of binning parameters that maintain spatial and frequency information and lead to desirable visual properties in the binned scatterplot.

>>>>

Binning algorithms used in making distributional approximations can be traced back to Pearson's work with the binomial approximation to the normal, where he mentions the need to define an origin and binwidth for segmenting the normal distribution \citep{Pearson1895}. \ktm{Sturges followed with an early work in formalizing histograms specification \citep{sturges1926choice}.}  More recently Scott has presented discussion on the importance of binning specification in the creation of histograms to appropriately display one dimensional density approximations \citep{scott1979}. \citet{scott1992} extends to the properties of multivariate binning strategies.

@@

When the individual points of a scatterplot are collapsed to bin centers to be displayed as tiles in a binned scatterplot there is a loss of the location information. \ktm{The tile size impacts the precision with which a user can read out location related attributes of the scatterplot; such as fine density patterns or outlier identification.} This loss can be expressed as the Euclidean distance between points and the visual center of the tiles (i.e.\ the bin centers). Note that other distance metrics could be used, but the Euclidean distance has a desirable interpretability in $\mathbb{R}^2$. The \textit{total spatial loss}, $L^S$, is defined as 
\begin{eqnarray}\label{totspatloss}
L^S = \sum_{i=1}^{n} L_i^S = \sum_{i=1}^{n} \sqrt{ \left(x_i-b_X(x_i)\right)^2 + \left(y_i-b_Y(y_i)\right)^2 }
\end{eqnarray} 
%
where $L_i^S$ is the loss in the  $i$th observation. 
%
Figure~\ref{fig:spatlossdemo2} visually displays the spatial loss for the data from  Table~\ref{DataReductionExample} as a result of standard rectangular binning. Observations $(x_i,y_i)$ and bin centers $\left(x_\ell^\ast,y_\ell^\ast\right)$ are displayed as black points and gray crosses, respectively. The length of line segments connecting these represent $L^S_i$, the spatial loss for each observation; thus, the combined length of all line segments represents the total spatial loss, $L^S$.  For random assignment the total spatial loss can be calculated by simply replacing the standard binning function with the random binning function in Equation~\ref{totspatloss}.

@@

% It is important to note that there is no spatial loss for the traditional scatterplot, as the points are rendered at exact coordinates. The reason we are willing to sacrifice precision of location information is to alleviate the inherent loss of frequency information from over-plotting, which can be immense for large data sets. When a point is rendered in a traditional scatterplot there is no graphical change once any additional points are placed at the same coordinates; thus visually implying a frequency of one data value at the location. This implicit loss of frequency information is then exacerbated when points are rendered as a circle with a radius large enough to also partially cover nearby points. In binned scatterplots, we have traded the exact locations for exact frequencies within bins. 
% 
Bin counts can be displayed using a continuous shading scale in HSV color space \citep{colorperception} and thus we can theoretically map frequency to shade perfectly. While the tiles of the binned scatterplot would be \textit{rendered} precisely, the ability of a human with average vision to extract that information by visually mapping the tile shade to a frequency scale in the plot legend is largely imprecise. Color perception of shade is \ktm{influenced by the contrasting effect of surrounding shades}, allowing an inaccurate mapping of tile shade to the corresponding shade in the plot legend \ktm{\citep{bartleson1967brightness,adelson1993perceptual,fairchild2013color}}. See Figure~\ref{fig:FreqLossDemoPlot} for a demonstration of context sensitivity of colors. It is therefore not realistic, to expect readers to be able to decode frequency \ktm{or accurately compare bin frequencies} from a continuous shade scheme, even though theoretically we can perceive shades continuously \citep{Leong}.  %Even though there is a lot of spread between the colors in this example, most onlookers have trouble answering the question about the relative relationship of the shading of the darkest/lightest points in each cloud of points.

@@

%Density changes are impossible to make out in the solid black areas of the plots. 97.4\%, 98.7\% and 99.1\% of the points are completely hidden behind at least one other point due to over-plotting in the coarse data for the uniform, normal and exponential simulations, respectively.
%
The \textit{data resolution} is defined as the smallest increment between successive data values. To observe loss from binning under more realistic conditions, we create three data sets by rounding the values from the originally simulated data sets to the nearest even number, thus a data resolution of 2 units in each dimension. This coarse version of the original simulated data is displayed in the bottom row of Figure~\ref{fig:SimDataScatterplots}. By exploring the loss properties for both the \textit{fine} and \textit{coarse} versions of the data, we identify which binning options are robust to the resolution at which data is recorded.

@@

To demonstrate the importance of properly selecting binwidths we consider the coarse version of the simulated bivariate-uniform data which is recorded to a resolution of two units in each dimension \ktm{-- thus even integer sized bins are most desirable. Figure~\ref{fig:UniformStripes}(a) displays the binned scatterplots under several scenarios. Under standard binning we see the white-space gaps with one-by-one unit bins, vertical and horizontal artificial stripes with five-by-five unit bins, and the appropriate view of an evenly spread density with four-by-four unit bins. Figure~\ref{fig:UniformStripes}(b) demonstrates that with coarse data resolution there are departures from the linear relationship between spatial loss and bin sizes; with small spikes in spatial loss where the bin size is highly misaligned with the data resolution and slight drops in spatial loss when aligned perfectly. In this we can see that artificial striping is a symptom of poorly aligned bin specification.}

@@

Note that random binning is effective at smoothing out the artificial striping patterns \ktm{when many point fell along bin boundaries -- accomplishing the motivation for its development. However due to inflated spatial loss due to the stochastic bin assignment, standard binning with properly selected bin dimension is the better option}.

@@

We now revisit the baseball data used earlier and construct binned scatterplots using the loss-based specification principles developed in the previous sections. \ktm{In the spatial binning step we learned three rules of thumb: (1) use smaller bins to achieve lower loss of spatial information, (2) standard binning has less loss than random binning but we need to be careful to choose bin dimensions that align with data resolution, and (3) offset the origin by a factor of half the data resolution to better center points within bins.}

@@

\ktm{Due to imperfect human perception, we may elect to use frequency binning to discretize the shading scale with the goal to make the loss of frequency information more explicit. If we aim to have highly accurate perceptual mapping,} it is recommended to use at most between four and seven distinct shades. This is done to minimize the frequency loss within the bounds of human perceptual ability to distinguish multiple shades simultaneously. Using quantile frequency binning or standard log frequency binning are shown to be reasonable methods -- with slight differences of interpretability -- for handling situations with heavily skewed bin counts. \ktm{Frequency binning to discretize the shading scale allows us to explicitly quantify the loss of frequency information, however future perception research is needed to quantify the implicit frequency loss in reading binned scatterplots using continuous shading and the impact of contrasting effects.}

Future implementations of software \ktm{should consider these loss-related findings in graphical tools for constructing binned scatterplots. This research provides suggestions for reasonable default settings of binning parameters that maintain spatial and frequency information and lead to desirable visual properties in the binned scatterplot.}
